---
title: "Time Series - Assignment 1"
author: "Damian Baeza"
date: "14 February 2018"
output: pdf_document
---

# Question 1

```{r}
require(ggplot2)
require(dplyr)
require(zoo)
require(xts)

load("UKGDP.RData")

yq <- paste(UKGDP$V1, UKGDP$V2)
df.yq <- cbind(UKGDP, as.yearqtr(yq))
df.xts <- xts(df.yq$V3, order.by = df.yq$`as.yearqtr(yq)`)

plot(df.xts, main = "UK GDP [£millions]")

```

As can be observed on the above plot, the gross domestic product of the UK has an trend that varies according to t (the year). This implies that the UK GDP does not have a constant trend, therefore the expected value of $X_{t}$ (UK GDP) is not constant. The aforementioned indicates that using an ARMA(p,q) model would not be appropiate.

# Question 2

```{r}
ARIMA1.Order <- c(0, 1, 1)
ARIMA2.Order <- c(1, 1, 0)
ARIMA3.Order <- c(1, 1, 1)

ARIMA1 <- arima(x = df.xts, order = ARIMA1.Order)
ARIMA2 <- arima(x = df.xts, order = ARIMA2.Order)
ARIMA3 <- arima(x = df.xts, order = ARIMA3.Order)

AICs <- c(ARIMA1$aic, ARIMA2$aic, ARIMA3$aic)

AIC.Probs <- exp((min(AICs) - AICs)/2)
```

## (a) Model selection

The ARIMA(1,1,1) model presents the lowest AIC value and further to that fact, the ARIMA(0,1,1) and ARIMA(1,1,0) models are $2.261 \cdot 10^{-13}$ and $1.707 \cdot 10^{-5}$ times as probable as the ARIMA(1,1,1) model to minimise the information loss.

The correlograms of the residuals are depicted in the following plots.

```{r}
# Correlogram of ARIMA(0,1,1) model
acf(ARIMA1$residuals)
# Correlogram of ARIMA(1,1,0) model
acf(ARIMA2$residuals)
# Correlogram of ARIMA(1,1,1) model
acf(ARIMA3$residuals)

# PACF of ARIMA(0,1,1) model
pacf(ARIMA1$residuals)
# PACF of ARIMA(1,1,0) model
pacf(ARIMA2$residuals)
# PACF of ARIMA(1,1,1) model
pacf(ARIMA3$residuals)

#Ljung-Box Test Statistics for lag = 1:50

lb <- function(lag, x, type = "Ljung-Box"){
  Box.test(x = x, lag = lag, type = type)$p.value
}

Lags <- 1:50
n.Lags <- length(Lags)

Model.p.values <- data.frame(ARIMA1 = numeric(n.Lags),
                             ARIMA2 = numeric(n.Lags),
                             ARIMA3 = numeric(n.Lags),
                             lag = Lags)

Model.p.values$ARIMA1 <- sapply(Lags, lb, x = ARIMA1$residuals)
Model.p.values$ARIMA2 <- sapply(Lags, lb, x = ARIMA2$residuals)
Model.p.values$ARIMA3 <- sapply(Lags, lb, x = ARIMA3$residuals)

# ggplot(data = Model.p.values, aes(x = Lags)) +
#   geom_line()
```
From the above correlograms it is possible to visualize that the residuals are uncorrelated for the ARIMA(1,1,1) model. This indicates that the estimated residuals are white noise.

## (b) Chosen model

The fitted ARIMA(1,1,1) model can be written as follows:

$$X_{t} = \alpha_{1}X_{t-1} + Z_{t} - \theta_{1}Z_{t-1}$$
